Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-05-10 papers/histogram/sad.tex
edits of sad paper

\begin{abstract}
  We present a novel Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a dynamic update factor (SAD-$\gamma$ or SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this MC method on the square-well fluid and
  compare the convergence time and average entropy error for a variety of weight-based
  Monte-Carlo methods. SAD rapidly converges to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.
\end{abstract}

@@

Liang began to consider that WL could be considered
a special case of the Stochastic Approximation approach, whose
convergence could be
mathematically proven~\cite{liang2006theory}. In
2007, Liang, Liu, and Carrol et al. argued that WL can be
considered a form of Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}. \davidsays{I'm here} While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

In this work, we have developed a novel algorithm based on SAMC that
does not require an array of non-physical, user-defined inputs and
therefore should be easily applicable to a given system. The method
does require the user to define a temperature range ($T_\text{min}$ to $T_\infty$).
We call this method SAD (Stochastic Approximation with a dynamic update
factor), and will discuss it in detail in the methods section. We
compare it along with several flat histogram methods which include WL,
$1/t$-WL, and SAMC.

We consider the square-well fluid i.e. a system of particles whose
interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is the simplest model
that ensures both attractive and repulsive forces are experienced by
interacting particles~\cite{barker1967-SW-perturbation, vega1992phase}.
The potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth.

@@

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are discussed and simulated for the
square-well fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Stochastic Approximation with a
dynamic $\gamma$ (SAD).

@@

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change the system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move and in what additional statistics must be collected in
order decide on that probability.

Flat histogram methods calculate
the density of states $D(E)$ for a discrete set of
energies~\cite{haber2018performance}. For this reason, energy binning
becomes an important consideration for systems with a continuum of
energies possible.
Energy bins are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Some methods such as
AdaWL~\cite{koh2013dynamically}
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies.  The method
introduced in this paper is designed to scale appropriately as bin size
is changed, but we do not test this scaling, as we use a system with
discrete energy levels.

In this section we will introduce four related flat histogram methods
which each rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.
A set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used
in this work.  Since $D(E) = S(E)$, the logarithm of the weights
can be thought of as an approximation of the entropy.

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreases as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how they schedule the
decreasing of $\gamma$.

@@

Wang-Landau's approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of steps at each energy during
each stage in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor, which is usually
$\frac12$.  The flatness is defined by the ratio between the minimum
value of the histogram and its average value.  When this flatness
reaches a specified threshold (typically 0.8), the $\gamma$ value is
decreased.  This approach requires that the energy range of interest
be known in advance, and difficulties can occur with this flatness
criteria due to the fact that some energies in this energy range might
never be sampled~\cite{haber2014transition}.  The entire process is
repeated until $\gamma$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or more
specifically, a set of energies) must be supplied, so that the
flatness criterion can be defined.

@@

\subsection{$1/t$-Wang-Landau}
The Wang-Landau algorithm's error saturation can be corrected by
modifying the update factor such that it does not decrease too quickly.
Belardinelli and Pereyra raised the question as to whether the update
factor should be decrease by $\nicefrac12$ or some other factor for
each update~\cite{belardinelli2007fast}. They implemented a schedule
that enforced that if the update factor $\gamma$ is less than
$\nicefrac{N_S}{t}$ then the update factor is set to
$\nicefrac{N_S}{t}$ where $N_S$ is the number of energy states and the
histogram is no longer tracked. Employing this scheduler, they found
that the error saturation is avoided since the correct density of
states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outlines minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_E}{t}$ scaling and switching from standard WL to
$1/t$-WL when the update factor $\gamma <
\nicefrac{N_E}{t}$~\cite{schneider2017convergence}. As per the original
$1/t$ implementation~\cite{belardinelli2007fast}, all energy states $N_E$
are required to be visited at least once $H(E) \neq 0$ effectively
avoiding the concept of `flatness'.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  In addition, the energy range need not be known
a priori.  The time to converge depends only on the choice of parameter
$t_0$.  Unfortunately, there is no prescription for finding an
acceptable value for $t_0$, and while the algorithm formally converges,
for a poor choice of $t_0$ that convergence can be far too slow to be
practical. Liang et al. give a rule of thumb in which $t_0$ is chosen
in the range from $2N_S$ to $100N_S$ where $N_S$ is the number of
energy bins~\cite{liang2007stochastic}.  Schneider et al. found that
for the Ising model this heuristic is helpful for small spin systems,
but that larger systems require an even higher $t_0$
value~\cite{schneider2017convergence}.  Also we find that for one case
with a small displacement distance for the square-well fluid, $t_0$
needs to be as much as two orders of magnitude higher than the rule of
thumb of $100N_S$ in order to converge in $10^{12}$ moves.

@@

\subsection{SAMC Convergence}\label{sec:samc-convergence}
A primary difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy of the system.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$, the lowest temperature of interest,
which is an immediate disadvantage of the method. However,
setting a physical parameter such as a minimum temperature
$T_\text{min}$ is possibly easier for a user
than determining in advance either an energy range or some unphysical
parameter $t_0$.

@@

While for SAMC, the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by $E_H -E_L$ where $E_H$ and
$E_L$ are defined below. The entropy numerator of the update factor in
general should scale with the total number of interesting energy states
$N_S$, since updates to the weights are distributed between that many
energy states.  The product $N_S\langle S\rangle$ is the total change
of entropy required (starting from constant weights) to find the true
entropy, and puts a lower bound on the convergence time. After
\emph{long} times, when all the energies have been a long time ago, we
wish for a lower update factor in order to more rapidly refine the
remaining error in entropy.  We track the time at which we first visted
each possible energy.  We define $t_L$ to be the last time that we
encountered an energy that we currently believe is in the energy range
of interest, so a $t\gg t_L$ we feel confident that we have established
the true energy range of interest. We gradually transition to a lower
update factor (but still asymptotically scaling as $\gamma_t \propto
1/t$ to ensure eventual convergence).  Finally, we wish for an update
factor that is \emph{never} greater than 1, because a very large update
factor could introduce errors in entropy that may take many iterations
to remove.  The SAD expression for $\gamma_t$ which incorporates these
ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n256}
  \caption{The update factor $\gamma_t$ for $N=256$ versus
    iteration for four different methods: WL, $1/t$-WL, SAMC, and SAD}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related methods
SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains constant
before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for many iterations,
and then decreases very rapidly, with $1/t$-WL behaving similarly before
transitioning to a $1/t$ behavior.
The update factor for SAD fluctuates
dynamically around a value less than 1 for early MC moves, and then
decreases as approximately $1/t$ while continuing to fluctuate as
new energies are found to be important.  At late times, the SAD $\gamma_t$
decreases as approximately $1/t^2$ until asymptoting to $N_S/t$, which
is the same as $1/t$-WL.

@@

During the simulation when considering a move inside of the energy
range $E_L \le E \le E_H$, Boltzmann weights are used.  If $E\ge E_H$,
the weight is taken to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.
This choice ensures that
if the maximum in entropy is at an energy $E_{\max}>E_H$, then the energy
$E_{\max}$
will eventually have the highest number of counts and the ratcheting will
result in $E_H\ge E_{\max}$ .
At lower energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.

@@

A significant advantage of SAD over other flat histogram methods is
that energy binning is handled dynamically. SAD should perform
independently for a reasonable range of binning because $\gamma \propto
\nicefrac{N_S}{t}$.  As the number of energy states $N_S$ found
increases (fine binning), the time spent $t$ in each bin will decrease
with the effect that the convergence should be roughly independent of
the bin size chosen.
This argument explains why SAD should perform well for continuum systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way~\cite{werlich2015stochastic} but this adds yet another parameter for the user to choose.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The number of important
               energies is $129$ of about $280$ energies.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area, and
           (b) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The number of important
               energies is $407$ of about $1800$ energies, and
           (d) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }\label{fig:fast-slow-gamma}
\end{figure*}

For the square-well fluid, our simulations treat two systems.  The first is
a smaller simulation with a particle number
of 50, a well-width of $\lambda = 1.3$, and a volume
corresponding to a filling fraction of $\eta = 0.3$. The second system
is larger, with a particle number
of 256, a well-width of $\lambda = 1.5$, and a volume
corresponding to a filling fraction of $\eta = 0.17$.
For each system we use a reasonable displacement distance $\delta_0 = 0.05\sigma$,
and for the smaller system we also use an unreasonably small displacement
distance of $0.005\sigma$. The simulations explore the
energy space of the systems with minimum reduced temperatures of
$T_{\text{min}} = 1/3$ for simulations of the smaller system, and $T_{\min}=1$
for the larger system.
All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

>>>>

\begin{abstract}
  \davidsays{Let's look at the abstract together.}
  We present a novel Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a dynamic update factor (SAD-$\gamma$ or SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this MC method on the square-well fluid and
  compare the convergence time and average entropy error for a variety of weight-based
  Monte-Carlo methods. SAD rapidly converges to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.
\end{abstract}

@@

Liang began to consider that WL could be considered a special case of
the Stochastic Approximation approach, whose convergence could be
mathematically proven~\cite{liang2006theory}. In 2007, Liang, Liu, and
Carrol \emph{et al.} developed Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}, and proved its convergence,
although the method still has a system specific user-defined parameter
which must be tuned when applying this algorithm to a new system.

In this work, we have developed an improved algorithm based on SAMC
that does not require an array of non-physical, user-defined inputs
and therefore should be easily applicable to any system. The method
does require the user to define a temperature range of ($T_\text{min}$
to $T_\infty$).  We call this method SAD (Stochastic Approximation
with a Dynamic update factor), and will discuss it in detail in the
methods section. We compare its convergence properties with three
existing flat histogram methods: WL, $1/t$-WL, and SAMC.

As a test case, we consider the square-well fluid i.e. a system of
particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth.
This model has the further advantage that because the energy is
discrete, binning is not required.

@@

In this work, we compare four flat histogram methods.  We outline the
general workings of each algorithm that we developed in detail while
summarizing algorithms that were developed in other works.  The
following methods are discussed and simulated for the square-well
fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and SAD.

@@

All the flat histogram Monte Carlo methods begin with randomly chosen
``moves'' which change the state of the system and must satisfy
detailed balance.  Each algorithm differs in how it determines the
probability of accepting a move and in what additional statistics must
be collected in order decide on that probability.

Flat histogram methods calculate the density of states $D(E)$ for a
discrete set of energies~\cite{haber2018performance}. For this reason,
energy binning becomes an important consideration for systems with a
continuum of possible energies.  Energy bins are typically of uniform
size for the entire energy continuum~\cite{fasnacht2004adaptive}. Some
methods such as AdaWL~\cite{koh2013dynamically} employ a tunable
mechanism for controlling the binning for low entropic states in order
to ensure the exploration of all energies.  The method introduced in
this paper is designed to scale appropriately as bin size is changed,
but we do not test this scaling, as we use a system with discrete
energy levels.

In this section we will introduce four closely related flat histogram
methods which each rely on a weight function $w(E)$.  In these
algorithms, the probability of accepting a move is given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
which biases the simulation in favor of energies with low weights.  A
set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus flat
histogram is a criteria for convergence for these methods.  To avoid
overflow error, since the weights may vary over more than a few
hundred orders of magnitude, the natural logarithm of the weights are
stored.  Since in the microcanonical ensemble the entropy is defined
as $S(E) \equiv k_B\ln(D(E))$, the logarithm of the weights can be
thought of as an approximation of the entropy.

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma_t
\end{equation}
where $t$ is number of the current move, $\gamma(t)$ is a
move-dependent update factor, and $E$ is the current energy.  This
update causes the random walk to avoid energies that have been
frequently sampled, leading to a rapid exploration of energy space.
This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreases as we decrease $\gamma_t$.  The four methods
differ primarily in how they schedule the decreasing of $\gamma_t$.

@@

The Wang-Landau approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of moves ending at each
energy in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor of $\frac12$.  The
flatness is defined by the ratio between the minimum value of the
histogram and its average value.  When this flatness reaches a
specified threshold (typically 0.8), the $\gamma$ value is decreased
and the histogram is reset to zero.  This approach requires that the
energy range of interest be known in advance, and difficulties can
occur with this flatness criteria due to the fact that some energies
in this energy range might never be
sampled~\cite{haber2014transition}.  The entire process is repeated
until $\gamma$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or in
general, a \emph{set} of energies) must be supplied, so that the
flatness criterion can be defined.

@@

\subsection{$1/t$-Wang-Landau}
The error saturation of Wang-Landau algorithm algorithm can be
corrected by modifying the update factor such that it does not
decrease too quickly.  Belardinelli and Pereyra raised the question as
to whether the update factor should be decrease by $\nicefrac12$ or
some other factor for each update~\cite{belardinelli2007fast}. They
implemented a schedule that enforced that if the update factor
$\gamma$ is ever less than $\nicefrac{N_S}{t}$ then the update factor
is set to $\nicefrac{N_S}{t}$ where $N_S$ is the number of energy
states and the histogram is no longer tracked. Employing this
scheduler, they found that the error saturation is avoided since the
correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outlines minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL
to $1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. \davidsays{Above
  we say that 1/t-WL does this.  One must be wrong?} As per the
original $1/t$ implementation~\cite{belardinelli2007fast}, the update
factor is decreased once all $N_S$ energy states are visited at least
once (i.e. $H(E) \neq 0$) effectively avoiding the concept of
`flatness'.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity, combined with is proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no prescription for finding
an acceptable value for $t_0$, and while the algorithm formally
converges, for a poor choice of $t_0$ that convergence can be far too
slow to be practical. Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.  We will describe below
one case we examined, in which $t_0$ needs to be as much as two orders
of magnitude higher than the rule of thumb of $100N_S$ in order to
converge in $10^{12}$ moves.

@@

\subsection{SAMC convergence time}\label{sec:samc-convergence}
A primary difficulty in using the SAMC method lies in identifying an
appropriate value for $t_0$.  Although SAMC is proven to formally
converge regardless of the $t_0$ value, a choice that is either too
high or too low will result in prohibitively slow convergence to the true
entropy of the system.  It is instructive to consider separately
values of $t_0$ that are too low or too high.

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC Algorithm that attempts to dynamically choose
the modification factor rather than relying on system dependent
parameters such as $t_0$ or $\gamma_0$.  There is an immediate
advantage of such an algorithm where parameters are chosen independent
of system size or type. Each flat-histogram method has unique
advantages and disadvantages.  Wang-Landau and $1/t$-WL require an
energy range for initialization.  SAMC removes this energy range
requirement but requires simulating every possible energy. Our
proposed method SAD requires the user to input $T_\text{min}$, the
lowest temperature of interest, which is an immediate disadvantage of
the method. However, identifying a minimum temperature of interest
$T_\text{min}$ may be easier for a user than determining in advance
an energy rang of intereste or the unphysical parameter $t_0$.

@@

While for SAMC, the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by $E_H -E_L$ where $E_H$ and
$E_L$ are defined below. The entropy numerator of the update factor in
general should scale with the total number of interesting energy states
$N_S$, since updates to the weights are distributed between that many
energy states.  The product $N_S\langle S\rangle$ is the total change
of entropy required (starting from constant weights) to find the true
entropy, and puts a lower bound on the convergence time. After
\emph{long} times, when all the energies have been a long time ago, we
wish for a lower update factor in order to more rapidly refine the
remaining error in entropy.  We track the time at which we first visted
each possible energy.  We define $t_L$ to be the last time that we
encountered an energy that we currently believe is in the energy range
of interest, so a $t\gg t_L$ we feel confident that we have established
the true energy range of interest. We gradually transition to a lower
update factor (but still asymptotically scaling as $\gamma_t \propto
1/t$ to ensure eventual convergence).  Finally, we wish for an update
factor that is \emph{never} greater than 1, because a very large update
factor could introduce very large errors in entropy that may take many iterations
to remove.  The SAD expression for $\gamma_t$ which incorporates these
ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
where $E_H$ and $E_L$ are the current estimates for the highest and
lowest energies of interest as defined below.  This factor
asymptotically has the same $1/t$ behavior as the original SAMC
algorithm and with the same $N_S$ prefactor used by the $1/t$-WL
method; however for earlier values of $t$, the update factor drops as
$1/t^2$ and jumps every time a new energy is determined to be of
interest.  This behavior allows SAD to dynamically prevent the update
factor from decreasing too rapidly.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n256}
  \caption{The update factor $\gamma_t$ versus iteration for $N=256$
    square-well atoms, for four different methods: WL, $1/t$-WL, SAMC,
    and SAD}
    \label{fig:gamma-vs-t}
\end{figure}
Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for the related
methods SAD, WL, $1/t$-WL, and SAMC.  For SAMC, $\gamma_t$ remains
constant before dropping as $1/t$.  WL $\gamma_t$ remains at 1 for
many iterations, and then decreases very rapidly, with $1/t$-WL
behaving similarly but decreasing more aggressively before
transitioning to a more conservative $1/t$ behavior.  The update
factor for SAD fluctuates dynamically around a value less than 1 for
early MC moves, and then decreases as approximately $1/t$ while
continuing to fluctuate as new energies are found to be important.  At
intermediate times, the SAD $\gamma_t$ decreases as $1/t^2$ before
asymptoting to $N_S/t$, which is the same as $1/t$-WL.

@@

During the simulation when considering a move inside of the energy
range of interest $E_L \le E \le E_H$, the weights are used as in the
three methods already described.  If $E\ge E_H$, the weight is taken
to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.  This choice ensures
that if the maximum in entropy is at an energy $E_{\max}>E_H$, then
the energy $E_{\max}$ will eventually have the highest number of
counts and the ratcheting will result in $E_H\ge E_{\max}$ .  At lower
energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.

@@

A significant advantage of SAD over SAMC---which the $1/t$-WL and WL
methods share \emph{after} they have found all the energies---is that
the schedule for $\gamma$ automatically responds to the choice of bin
size. SAD should perform similarly over a wide range of bin sizes
because $\gamma \propto \nicefrac{N_S}{t}$.  As the number of energy
states $N_S$ found increases (fine binning), the time spent $t$ in
each bin will decrease with the effect that the convergence should be
roughly independent of the bin size chosen.  SAMC could be used with a
prefactor $\gamma_0$ to aid in a similar
way~\cite{werlich2015stochastic} but this adds yet another parameter
for the user to choose.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area, and
           (b) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The average entropy error for each MC method for the
               same physical system with a smaller displacement distance
               $\delta_0 = 0.005\sigma$,
               as a function of number of iterations run, and
           (d) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }\label{fig:fast-slow-gamma}
\end{figure*}

We tested the algorithms on two square-well fluid systems.  The first
is a smaller simulation with a particle number of 50, a well-width of
$\lambda = 1.3$, and a volume corresponding to a filling fraction
(defined as the fraction of volume filled by atoms) of $\eta =
0.3$. The second system is larger, with a particle number of 256, a
well-width of $\lambda = 1.5$, and a volume corresponding to a filling
fraction of $\eta = 0.17$.  For each system we use a reasonable
root-mean-square displacement distance $\delta_0 = 0.05\sigma$ for
proposed moves, and for the smaller system we also use an unreasonably
small displacement distance of $0.005\sigma$. The simulations explore
the energy space of the systems with minimum reduced temperatures of
$T_{\text{min}} = 1/3$ for simulations of the smaller system, and
$T_{\min}=1$ for the larger system.  All simulations lead to the
minimum important energy $E_{\min}$ and maximum entropy energy
$E_{\max}$ being calculated (with the exception of the WL methods
where both of these parameters are needed a priori).
